---
title: "Model Based Clustering using MClust"
author: "Anthony Kamau"
toc: yes
output: pdf_document
---

# Introduction
We shall apply model based clustering on a dataset of bank note measurements

Load the packages and data. The data is built into the `mclust` package.

```{r setup, warning=FALSE, message=FALSE}
# install.packages(c("mclust", "patchwork", "ggridges"))

# load packages
library(mclust)
library(tidyverse)
library(patchwork)
library(ggridges)

# load data
df <- as.tibble(banknote)
```

# Data Exploration
```{r}
# help file to understand dataset
?banknote

# scatter plot
df |>
  ggplot(aes(x = Left, y = Right, colour = Status)) +
  geom_point(position = position_jitter())

# The classes are not easy to distinguish, there is considerable overlap between 
# the two classes
```

```{r, message=FALSE}
# remove Status column from the data set
df <- df |> select(-Status)

# density plot for all columns in the data set
df |> ggplot(aes(x = Length)) + geom_density() + theme_minimal() +
  df |> ggplot(aes(x = Left)) + geom_density() + theme_minimal() +
  df |> ggplot(aes(x = Right)) + geom_density() + theme_minimal() +
  df |> ggplot(aes(x = Bottom)) + geom_density() + theme_minimal() +
  df |> ggplot(aes(x = Top)) + geom_density() + theme_minimal() +
  df |> ggplot(aes(x = Diagonal)) + geom_density() + theme_minimal()

# The diagonal feature looks good. Look at the two bumps in its density plot

# colorful alternative
df |> mutate(across(everything(), scale)) |> 
  pivot_longer(everything(), names_to = "feature", values_to = "value") |>
  ggplot(aes(x = value, y = feature, fill = feature)) + 
  geom_density_ridges() +
  scale_fill_viridis_d(guide = "none") +
  theme_minimal()

# 6 univariate density plots (Length, Left, Right, Bottom, Top, Diagonal) are 
# compared to see which single feature shows the clearest separation into 
# two groups (bimodal meaning 2 distinct bumps)
# Diagonal stands out with two clearly separated peaks in its density curve,
# while the others are mostly single-peaked (unimodal) or only weakly bimodal.
# Diagonal the best single candidate feature for clustering, since a two-humped 
# density strongly suggests two underlying subpopulations (exactly what model-based 
# clustering is designed to detect.
```
# Univariate model-based clustering
Perform model-based clustering with 2 clusters on the `Diagonal` feature, assuming equal 
variance i.e. forcing both clusters to share the same variance
```{r}
fit_E_2 <- Mclust(df$Diagonal, G = 2, modelNames = "E")

# means
fit_E_2$parameters$mean

# variance
fit_E_2$parameters$variance$sigmasq

# Means (139.45 and 141.52), two distinct centre, consistent with the two 
# humped density density seen earlier
# Variance (0.244), shared by both clusters, since equal clusters was assumed
```
Compute the model's BIC using its log-likelihood and compare it to the BIC stored
in the model object.
```{r}
# This model has 4 parameters (1 class probability pi, 2 means, and 1 variance)
# BIC computed 
- 2 * fit_E_2$loglik + log(nrow(df)) * 4

# BIC from mclust package
fit_E_2$bic

# BIC from the Mclust package is wrongly multiplied by -1

# manual BIC calculation using the formula (-2 × log-likelihood + log(n) × m, 
# where m = 4 parameters i.e. 1 mixing proportion, 2 means, 1 shared variance) 
# gives 569.47, matching fit_E_2$bic exactly (aside from a sign flip). 
# Mclust stores BIC as the negative of the conventional definition, 
# so fit_E_2$bic = -569.4667
# You need to negate it inorder to make comparissons
```


```{r}
# plot the density using base R plots
plot(fit_E_2, "density", xlab = "Diagonal measurement")
# add the observations using rug marks
rug(df$Diagonal)

# shows two symmetric, equally-wide bell curves centered near 139.5 and 141.5, 
# with rug marks below showing the actual raw data points. 
# The two bumps have the same width because equal variance was forced.
```

Perform model-based clustering with 2 clusters on the `Diagonal` feature again, but
now assume unequal variances.
```{r}
# Let's set the seed for reproducibility
fit_V_2 <- Mclust(df$Diagonal, G = 2, modelNames = "V")

# mean
fit_V_2$parameters$mean

# variance
fit_V_2$parameters$variance$sigmasq

# Means (139.50 and 141.56), nearly identical to the equal variance model
# Variance (0.359 and 0.150), now clearly different. The first cluster (lower Diagonal)
# is more spread out than the second
```
```{r}
# plot the density using base R plots
plot(fit_V_2, "density", xlab = "Diagonal measurement")
# add the observations using rug marks
rug(df$Diagonal)

# The left bump is wider/shorter (more spread, less peaked) and the right bump 
# is narrower/taller (more concentrated), directly reflecting the different variances.
```


```{r}
# This model has 5 parameters: 1 class probability pi, 2 means, and 2 variance
# BIC computed
-2*fit_V_2$loglik

# BIC from mclust package
fit_V_2$bic
```

## Deviance Comparisson
```{r}
# model with equal variance
-2*fit_E_2$loglik

# model with unequal variance
-2*fit_V_2$loglik

# Lower deviance means better fit to the data, so the V (unequal variance) model 
# fits better. 
# This makes sense, since it has an extra parameter giving it more flexibility 
# to match the data exactly.
```

## BIC Comparisson
`NB:` Remember, the package shows negative BIC
```{r}
# model with equal variance
-fit_E_2$bic

# model with unequal variance
-fit_V_2$bic

# The V model still wins on BIC (lower is better), but the gap between the two models
# has narrowed compared to the deviance comparison. 
# This is exactly what BIC is designed to do. It penalizes the extra parameter in
# the V model, so while V still fits better overall, its advantage is less 
# dramatic than raw deviance alone suggests. BIC is protecting against overfitting.
```

# Multivariate model base clustering
We will now use all the available information in the data set to cluster the
observations

```{r}
# mlcust with all 6 features
fit <- Mclust(df)

summary(fit)

# The optimal model is a VVE model with 3 components


# mean parameters
fit$parameters$mean

# 3 clusters times 6 variables = 18 mean parameters


# Running Mclust(df) with no constraints lets it search across all covariance 
# structures and 1–9 possible cluster counts, then pick whichever combination 
# minimizes BIC. The winner: a VVE model with 3 components, meaning:
# Variable volume (cluster can have different sizes/spread)
# Variable shape (clusters can be different shapes, not just circular)
# Equal orientation (all clusters point the same direction in feature space)

# 3 clusters, with 18, 98, and 84 observations respectively (200 total)
# Multivariate clustering found 3 clusters, not 2, even though the data only has 
# 2 true classes (genuine/counterfeit). 
# The data may genuinely split into 3 identifiable subgroups, even though our 
# external knowledge says there are only 2 real-world categories.
```
Forcing exactly 2 clusters (to match the known genuine/counterfeit split) with the most flexible
covariance structure (VVV = variable volume, variable shape, variable orientation). The resulting
matrix of bivariate density contour plots lets you inspect, for every pair of variables, 
how well-separated the two estimated clusters are.
```{r}
# 2-component VVV model 
VVV2 <- Mclust(df, 2, "VVV")

# matrix of bivariate contour (density) plots
plot(VVV2, "density")

# Diagonal-top, and diagonal-bottom separate the classes very well
# Diagonal-Top and Diagonal-Bottom pairs show the clearest separation 
# (two distinct, non-overlapping contour blobs), while most other pairs 
# (like Left-Right, matching the very first plot in the document) show heavily 
# overlapping contours, reinforcing that Diagonal is the single most discriminating 
# feature, and combining it with Top or Bottom gives especially clean separation.
```
Plotting Left vs Right again, but now coloring points by the model's estimated 
cluster assignment (not the true Status labels) and sizing points by classification 
uncertainty.
```{r}
df |> ggplot(aes(x = Left, y = Right, colour = as_factor(VVV2$classification),
                 size = VVV2$uncertainty)) + 
  geom_point(position = position_jitter()) +
  scale_size(range = c(1, 3)) +
  labs(colour = "Cluster", size = "Uncertainty")


# almost all points are small (low uncertainty) meaning the model is very confident 
# in most assignments, with only 3 points showing any meaningful uncertainty, 
# and notably, those uncertain points don't sit near the visual boundary between 
# the two colors in this particular Left/Right view.
# only 3 points are slightly uncertain, and they are not around the border of the 
# classes in these two dimensions.
# Apparently, the other dimensions give enough information
# about those points near the border
```

